---
title: "Data Cleaning and Imputation Pipeline Report"
subtitle: "Retail Store Sales Dataset"
author: "Onika Bevny"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  dpi = 300
)

library(readr)
library(dplyr)
library(knitr)
library(ggplot2)
library(flextable)
library(officer)
library(scales)

theme_set(theme_minimal(base_size = 11))

make_table <- function(data, caption = NULL) {
  flextable(data) |>
    set_caption(caption = caption) |>
    autofit() |>
    fontsize(size = 9, part = "all") |>
    align(align = "center", part = "all") |>
    theme_vanilla()
}
```

# Introduction

This report presents the design, implementation, and execution of a fully reproducible data cleaning and imputation pipeline developed for the `retail_store_sales` dataset.

The pipeline was created in R using a structured and modular workflow that incorporated multiple stages, including data ingestion, diagnostics, cleaning, imputation, validation, export generation, and automated reporting.

The primary objective of the pipeline was to identify, assess, and address data quality issues while maintaining transparency, consistency, and reproducibility throughout the entire analytical process.

Key areas addressed within the pipeline included:

-   Detection and handling of missing values
-   Identification of cross-field inconsistencies
-   Validation of important numeric relationships
-   Recovery of logically recoverable values
-   Application of basic imputation techniques
-   Generation of reproducible multi-format outputs
-   Automated creation of analytical summaries and reports

------------------------------------------------------------------------

# Dataset Overview

The dataset assigned for this project was the `retail_store_sales` dataset. The analysis and pipeline development were limited to a predefined set of variables provided within the assignment scope.

The variables included in the project were:

-   Transaction ID
-   Customer ID
-   Category
-   Item
-   Quantity
-   Price Per Unit
-   Total Spent
-   Payment Method
-   Location
-   Transaction Date
-   Discount Applied

The pipeline was developed as a structured end-to-end workflow designed to process the dataset systematically while maintaining reproducibility and consistency across all stages of the project.

The workflow consisted of the following stages:

1.  Data ingestion
2.  Data diagnostics
3.  Data cleaning
4.  Missing value imputation
5.  Validation and consistency checks
6.  Data dictionary generation
7.  Multi-format dataset export
8.  Automated report generation

Each stage was implemented as an independent script within the pipeline to support modularity, traceability, and ease of maintenance.

```{r diagnostics-overview}
diagnostics_overview <- read_csv(
  "outputs/diagnostics_overview.csv",
  show_col_types = FALSE
)

make_table(
  diagnostics_overview,
  "Pipeline Diagnostics Overview"
)
```

------------------------------------------------------------------------

# Data Quality Diagnostics Before Cleaning

Initial diagnostics were conducted to evaluate the quality of the dataset prior to cleaning and transformation.

The diagnostics stage focused on:

-   Missingness analysis

-   Sentinel value detection

-   Duplicate checks

-   Date parsing validation

-   Numeric range validation

-   Spending logic validation

-   Cross-field recoverability assessment

```{r missing-summary-table}
missing_summary <- read_csv(
  "outputs/missing_summary.csv",
  show_col_types = FALSE
)

make_table(
  missing_summary,
  "Missingness Summary Before Cleaning"
)
```

## Missingness Visualization

```{r missingness-before-chart}
ggplot(
  missing_summary,
  aes(
    x = reorder(variable, missing_percent),
    y = missing_percent
  )
) +
  geom_col(fill = "#2C7FB8") +
  coord_flip() +
  labs(
    title = "Missingness Before Cleaning",
    x = "Variable",
    y = "Missing Percent"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.y = element_text(size = 8)
  )
```

------------------------------------------------------------------------

# Cleaning Decisions and Transformations

The cleaning stage focused on correcting recoverable inconsistencies while preserving the integrity of the original dataset.

## Cross-Field Recovery

Cross-field recovery was implemented using the deterministic relationship:

**Total Spent = Quantity × Price Per Unit**

Missing `price_per_unit` values were recovered where both `quantity` and `total_spent` were available.

Rows that could not be logically recovered were preserved for the imputation stage.

```{r missing-comparison-table}
missing_comparison <- read_csv(
  "outputs/missing_comparison.csv",
  show_col_types = FALSE
)

make_table(
  missing_comparison,
  "Missingness Before and After Cleaning"
)
```

## Recovery Visualization

```{r recovery-chart}
ggplot(
  missing_comparison,
  aes(
    x = reorder(variable, recovered),
    y = recovered
  )
) +
  geom_col(fill = "#2C7FB8") +
  coord_flip() +
  labs(
    title = "Values Recovered During Cleaning",
    x = "Variable",
    y = "Recovered Values"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.y = element_text(size = 8)
  )
```

------------------------------------------------------------------------

# Imputation Methods Applied

The imputation stage addressed unresolved missing values remaining after cleaning.

## Median Imputation

Median imputation was applied to the following numeric variables:

-   `quantity`
-   `total_spent`

Median values were calculated using observed non-missing data only.

## Hot-Deck Imputation

A simple category-based hot-deck approach was applied to the `item` variable.

Missing item values were replaced using the most frequently observed item within the same product category.

## Preserved Missingness

Missing values in `discount_applied` were intentionally retained because there was insufficient information to reliably determine whether a discount had been applied.

A missingness indicator variable was created to preserve this information for downstream analysis.

```{r imputation-comparison-table}
imputation_comparison <- read_csv(
  "outputs/missing_imputation_comparison.csv",
  show_col_types = FALSE
)

make_table(
  imputation_comparison,
  "Missingness Before and After Imputation"
)
```

## Imputation Visualization

```{r imputation-chart}
ggplot(
  imputation_comparison,
  aes(
    x = reorder(variable, imputed_or_resolved),
    y = imputed_or_resolved
  )
) +
  geom_col(fill = "#2C7FB8") +
  coord_flip() +
  labs(
    title = "Values Imputed or Resolved",
    x = "Variable",
    y = "Count"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.y = element_text(size = 8)
  )
```

------------------------------------------------------------------------

# Post-Cleaning and Post-Imputation Summary

The following section summarizes the final missingness structure of the dataset after all cleaning and imputation procedures were completed.

```{r final-missingness-table}
final_missingness <- read_csv(
  "outputs/final_missingness_summary.csv",
  show_col_types = FALSE
)

make_table(
  final_missingness,
  "Final Missingness Summary"
)
```

## Final Missingness Visualization

```{r final-missingness-chart}
ggplot(
  final_missingness,
  aes(
    x = reorder(variable, missing_percent_final),
    y = missing_percent_final
  )
) +
  geom_col(fill = "#2C7FB8") +
  coord_flip() +
  labs(
    title = "Final Missingness by Variable",
    x = "Variable",
    y = "Final Missing Percent"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.y = element_text(size = 8)
  )
```

------------------------------------------------------------------------

# Numeric Variable Distributions

The following visualizations summarize the distributions of key numeric variables in the finalized dataset.

```{r quantity-distribution}
final_data <- read_csv(
  "outputs/final_cleaned_imputed_data.csv",
  show_col_types = FALSE
)

ggplot(final_data, aes(x = quantity)) +
  geom_histogram(
    bins = 30,
    fill = "#41B6C4",
    color = "white"
  ) +
  labs(
    title = "Distribution of Quantity",
    x = "Quantity",
    y = "Count"
  ) +
  theme(
    plot.title = element_text(face = "bold")
  )
```

```{r total-spent-distribution}
ggplot(final_data, aes(x = total_spent)) +
  geom_histogram(
    bins = 30,
    fill = "#41B6C4",
    color = "white"
  ) +
  labs(
    title = "Distribution of Total Spent",
    x = "Total Spent",
    y = "Count"
  ) +
  theme(
    plot.title = element_text(face = "bold")
  )
```

------------------------------------------------------------------------

# Validation Summary

Final validation checks were conducted after imputation to ensure that cleaning and imputation procedures did not introduce inconsistencies or invalid values.

Validation focused on:

-   Remaining total spent mismatches
-   Invalid quantity values
-   Invalid price values
-   Invalid total spent values

```{r validation-summary-table}
validation_summary <- read_csv(
  "outputs/imputation_validation_summary.csv",
  show_col_types = FALSE
)

make_table(
  validation_summary,
  "Validation Summary After Imputation"
)
```

------------------------------------------------------------------------

# Limitations and Known Risks

Despite the successful implementation of the pipeline, several limitations should be considered:

-   Some missing values in `discount_applied` were intentionally retained because there was insufficient information to accurately determine whether a discount had originally been applied.
-   The use of median imputation for numeric variables may reduce the natural variability present within the dataset.
-   The hot-deck imputation approach assigned missing `item` values using the most frequently observed item within each category, which may not always reflect the original missing value.
-   More advanced predictive or machine learning-based imputation methods were not implemented in order to remain consistent with the scope and requirements of the assignment.

------------------------------------------------------------------------

# Conclusion

This project successfully developed a fully reproducible data cleaning and imputation pipeline for the assigned dataset. The pipeline automated the complete workflow, including data ingestion, diagnostics, cleaning, imputation, validation, documentation, export generation, and automated reporting.

Major outcomes of the project included:

-   Recovery of missing values through cross-field logical relationships
-   Application of multiple basic imputation techniques, including median and hot-deck imputation
-   Verification of numeric and logical consistency rules throughout the pipeline
-   Generation of automated diagnostics, summaries, and validation outputs
-   Export of finalized datasets into multiple reproducible formats
-   Automated generation of a structured Word report summarizing the full analytical workflow

Overall, the pipeline provides a transparent, organized, and reproducible framework for managing data quality issues while remaining aligned with the scope and requirements of the assignment.
