---
title: "Problem Inventory — Week 1"
author: "Onika Bevny"
date: today
format:
  docx:
    toc: true
    number-sections: true
---

## Overview

Initial diagnostics were conducted on the `retail_store_sales` dataset to identify data quality issues within the scoped variables.

The dataset contains **12,575 observations and 11 variables**, covering transaction details, customer information, pricing, quantities, and discount indicators.

This document summarizes the issues identified during the diagnostics stage and outlines areas requiring attention in the cleaning phase.

## Missing Values

Missing values were assessed across all variables using summary statistics.

```{r}
library(readr)
library(knitr)

missing_summary <- read_csv("../outputs/missing_summary.csv", show_col_types = FALSE)
kable(missing_summary)
```

### Key Findings

Missing values are present in variables such as: price_per_unit, quantity, total_spent and discount_applied

These missing values will be addressed during the cleaning stage, particularly where recovery is possible using cross-field relationships.

## Sentinel Values

The dataset was evaluated for sentinel values such as "ERROR", "UNKNOWN", and blank strings.

```{r}
sentinel_summary <- read_csv("../outputs/sentinel_summary.csv", show_col_types = FALSE)
kable(sentinel_summary)
```

### Findings

No sentinel values were detected across any variables.

This indicates that missing or invalid values are not encoded using placeholder strings, simplifying the cleaning process.

## Duplicate Records

Duplicate checks were performed on both transaction identifiers and full rows.

```{r}
diagnostics_overview <- read_csv("../outputs/diagnostics_overview.csv", show_col_types = FALSE)
kable(diagnostics_overview)
```

### Findings

Duplicate Transaction IDs: 0

Exact duplicate rows: 0

This suggests that each transaction record is unique and no duplicate removal is required.

## Date Validation

Transaction dates were assessed for parsing and validity.

### Findings 

Date parsing issues: 0

All transaction dates are correctly formatted and require no correction.

## Numeric and Logical Checks

Numeric variables (quantity, price_per_unit, total_spent) were assessed for invalid values.

### Findings

No negative or non-positive values detected No inconsistencies identified in numeric ranges

Additionally, the relationship:

Total Spent = Quantity × Price Per Unit

was validated.

### Findings

Spending mismatches: 0

This indicates strong internal consistency within the dataset.

## Cross-Field Recovery

A key finding from the diagnostics is the presence of 609 records where missing values can be recovered using relationships between variables.

Examples include:

Recovering price_per_unit when quantity and total_spent are available Recovering total_spent when quantity and price_per_unit are available

These cases will be handled during the cleaning stage where appropriate.

## Discount Applied Variable

The discount_applied variable was assessed for consistency.

```{r}
discount_summary <- read_csv("../outputs/discount_applied_summary.csv", show_col_types = FALSE)
kable(discount_summary)
```

### Findings

Values in this variable require standardization into a consistent format.

This will be addressed during the cleaning stage.

## Summary of Key Issues

Overall, the dataset is relatively clean, with:

No duplicate records No date parsing issues No numeric inconsistencies No sentinel values

The primary issue identified is:

Missing values that are largely recoverable through cross-field relationships.

## Planned Actions for Cleaning Stage

Based on the findings, the following actions will be taken:

Recover missing values using cross-field calculations where possible Standardize the discount_applied variable Ensure consistency across categorical variables Maintain detailed logs of all transformations
